---
title: "Trading Strategy Backtester"
subtitle: "Golden Cross vs RSI Mean Reversion vs 12-Month Momentum"
author: "Manchet Singh"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
    df_print: paged


params:
  ticker:    "AAPL"
  from:      "2015-01-01"
  to:        "2024-12-31"
  ma_fast:   50
  ma_slow:   200
  rsi_low:   30
  rsi_high:  70
  rsi_n:     14
  mom_days:  252
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

---

# Overview

This report backtests three systematic trading strategies on
**`r params$ticker`** from **`r params$from`** to **`r params$to`**
using free data from Yahoo Finance.

| Strategy | Type | Core idea |
|---|---|---|
| Golden Cross | Trend following | Buy when 50-day MA crosses above 200-day MA |
| RSI Mean Reversion | Contrarian | Buy when RSI < `r params$rsi_low`, sell when RSI > `r params$rsi_high` |
| 12-Month Momentum | Momentum | Stay long when 12-month return is positive |

> All charts are interactive — hover, click legend items to toggle,
> drag to zoom.

---

# Setup

## Load packages

```{r packages}
# Run once in your console if not yet installed:
# install.packages(c("quantmod","TTR","PerformanceAnalytics",
#                    "tidyverse","plotly","kableExtra","DT","xts"))

library(quantmod)
library(TTR)
library(PerformanceAnalytics)
library(tidyverse)
library(plotly)
library(kableExtra)
library(DT)
library(xts)
```

---

# Data

```{r data}
raw <- tryCatch({
  getSymbols(
    params$ticker,
    src         = "yahoo",
    from        = params$from,
    to          = params$to,
    auto.assign = FALSE,
    warnings    = FALSE
  )
}, error = function(e) {
  stop(paste("Could not download", params$ticker,
             "— check the ticker and date range."))
})

prices <- Ad(raw)
n_days <- nrow(prices)

cat("Ticker:       ", params$ticker, "\n")
cat("Trading days: ", n_days, "\n")
cat("From:         ", as.character(min(index(prices))), "\n")
cat("To:           ", as.character(max(index(prices))), "\n")
cat("MA check:     ",
    ifelse(n_days > params$ma_slow,
           "PASS", "FAIL — extend 'from' date"), "\n")
cat("Momentum check:",
    ifelse(n_days > params$mom_days,
           "PASS", "FAIL — extend 'from' date"), "\n")
```

## Price chart

```{r price-chart, fig.height=3.5}
plot_ly(
  x    = index(prices),
  y    = as.numeric(prices),
  type = "scatter",
  mode = "lines",
  name = params$ticker,
  line = list(color = "#2C2C2A", width = 1.2)
) %>%
  layout(
    title     = paste(params$ticker, "— Adjusted close price"),
    xaxis     = list(title = ""),
    yaxis     = list(title = "Price ($)", tickprefix = "$"),
    hovermode = "x unified"
  ) %>%
  config(displayModeBar = FALSE)
```

---

# Strategy 1 — Golden Cross

## How it works

- **Buy** when the `r params$ma_fast`-day MA crosses **above** the
  `r params$ma_slow`-day MA
- **Sell** (go to cash) when it crosses back **below**
- Signal lagged 1 day — no look-ahead bias

This is a **trend-following** strategy. It keeps you in the market
during sustained uptrends and pulls you out during prolonged downtrends.
It gets whipsawed in choppy, sideways markets.

```{r golden-cross}
golden_cross <- function(prices, fast = 50, slow = 200) {
  ma_fast   <- SMA(prices, n = fast)
  ma_slow   <- SMA(prices, n = slow)
  signal    <- ifelse(ma_fast > ma_slow, 1, 0)
  signal    <- lag(signal, 1)
  signal[is.na(signal)] <- 0
  daily_ret <- dailyReturn(prices)
  strat_ret <- daily_ret * signal
  list(
    signal    = signal,
    daily_ret = daily_ret,
    strat_ret = strat_ret,
    ma_fast   = ma_fast,
    ma_slow   = ma_slow
  )
}

gc_result <- golden_cross(prices, params$ma_fast, params$ma_slow)
cat("Days in market:",
    sum(as.numeric(gc_result$signal) == 1, na.rm = TRUE),
    "of", n_days, "\n")
```

## Price chart with moving averages

```{r gc-chart, fig.height=4}
plot_ly() %>%
  add_lines(
    x    = index(prices),
    y    = as.numeric(prices),
    name = "Price",
    line = list(color = "#2C2C2A", width = 1)
  ) %>%
  add_lines(
    x    = index(prices),
    y    = as.numeric(gc_result$ma_fast),
    name = paste0("MA", params$ma_fast, " fast"),
    line = list(color = "#185FA5", width = 2)
  ) %>%
  add_lines(
    x    = index(prices),
    y    = as.numeric(gc_result$ma_slow),
    name = paste0("MA", params$ma_slow, " slow"),
    line = list(color = "#D85A30", width = 2)
  ) %>%
  layout(
    title     = paste(params$ticker, "— Golden Cross moving averages"),
    xaxis     = list(title = ""),
    yaxis     = list(title = "Price ($)", tickprefix = "$"),
    hovermode = "x unified",
    legend    = list(orientation = "h", x = 0, y = -0.2)
  ) %>%
  config(displayModeBar = FALSE)
```

> When blue (fast MA) is above orange (slow MA) the strategy is long.
> When orange is above blue the strategy is in cash.

---

# Strategy 2 — RSI Mean Reversion

## How it works

RSI measures how hard and fast a stock has moved, on a 0–100 scale:

- **Buy** when RSI drops below `r params$rsi_low` — stock is oversold
- **Sell** when RSI rises above `r params$rsi_high` — stock is overbought

This is a **contrarian** strategy — the opposite of trend following.
It bets that extreme moves snap back toward the average. Works best
in choppy, range-bound markets. Struggles in strong trends.

```{r rsi-strategy}
rsi_strategy <- function(prices, rsi_low = 30, rsi_high = 70, n = 14) {
  rsi_vals <- RSI(prices, n = n)
  position <- rep(0, length(prices))
  in_trade <- FALSE
  for (i in (n + 1):length(prices)) {
    if (!in_trade && !is.na(rsi_vals[i]) &&
        rsi_vals[i] < rsi_low) {
      in_trade <- TRUE
    } else if (in_trade && !is.na(rsi_vals[i]) &&
               rsi_vals[i] > rsi_high) {
      in_trade <- FALSE
    }
    position[i] <- as.integer(in_trade)
  }
  signal    <- lag(xts(position, index(prices)), 1)
  signal[is.na(signal)] <- 0
  daily_ret <- dailyReturn(prices)
  strat_ret <- daily_ret * signal
  list(
    signal    = signal,
    daily_ret = daily_ret,
    strat_ret = strat_ret,
    rsi       = rsi_vals
  )
}

rsi_result <- rsi_strategy(
  prices,
  params$rsi_low,
  params$rsi_high,
  params$rsi_n
)
cat("Days in market:",
    sum(as.numeric(rsi_result$signal) == 1, na.rm = TRUE),
    "of", n_days, "\n")
```

## RSI indicator chart

```{r rsi-chart, fig.height=3.5}
plot_ly() %>%
  add_lines(
    x    = index(prices),
    y    = as.numeric(rsi_result$rsi),
    name = paste0("RSI (", params$rsi_n, "-day)"),
    line = list(color = "#534AB7", width = 1.5)
  ) %>%
  add_segments(
    x    = min(index(prices)), xend = max(index(prices)),
    y    = params$rsi_high,    yend = params$rsi_high,
    name = paste("Overbought —", params$rsi_high),
    line = list(color = "#D85A30", dash = "dash", width = 1.5)
  ) %>%
  add_segments(
    x    = min(index(prices)), xend = max(index(prices)),
    y    = params$rsi_low,     yend = params$rsi_low,
    name = paste("Oversold —", params$rsi_low),
    line = list(color = "#185FA5", dash = "dash", width = 1.5)
  ) %>%
  layout(
    title     = paste(params$ticker, "— RSI indicator"),
    xaxis     = list(title = ""),
    yaxis     = list(title = "RSI", range = c(0, 100)),
    hovermode = "x unified",
    legend    = list(orientation = "h", x = 0, y = -0.2)
  ) %>%
  config(displayModeBar = FALSE)
```

---

# Strategy 3 — 12-Month Momentum

## How it works

Momentum is one of the most researched anomalies in finance, first
formally documented by Jegadeesh & Titman (1993). The core idea is
simple: **stocks that have gone up over the past year tend to keep
going up.**

The rule used here:

- Calculate the **12-month return** (252 trading days) for the stock
  on every day
- **Stay long** as long as the 12-month return is **positive** —
  momentum is intact
- **Go to cash** when the 12-month return turns **negative** —
  momentum has broken down
- Signal lagged 1 day — no look-ahead bias

This is different from Golden Cross because it compares the stock to
its own price from a year ago rather than comparing two moving
averages. It is a slower, longer-term signal that generates far
fewer trades.

```{r momentum}
momentum_strategy <- function(prices, n = 252) {

  # 12-month return: today's price vs price n days ago
  mom_ret <- ROC(prices, n = n, type = "discrete")

  # Signal: 1 = in market (positive momentum), 0 = cash
  signal  <- ifelse(mom_ret > 0, 1, 0)
  signal  <- lag(signal, 1)
  signal[is.na(signal)] <- 0

  daily_ret <- dailyReturn(prices)
  strat_ret <- daily_ret * signal

  list(
    signal    = signal,
    daily_ret = daily_ret,
    strat_ret = strat_ret,
    mom_ret   = mom_ret
  )
}

mom_result <- momentum_strategy(prices, params$mom_days)
cat("Days in market:",
    sum(as.numeric(mom_result$signal) == 1, na.rm = TRUE),
    "of", n_days, "\n")
```

## 12-month rolling return chart

```{r mom-chart, fig.height=4}
mom_pct <- as.numeric(mom_result$mom_ret) * 100

plot_ly() %>%
  add_lines(
    x    = index(prices),
    y    = mom_pct,
    name = "12-month return",
    line = list(color = "#534AB7", width = 1.5),
    fill = "tozeroy",
    fillcolor = ifelse(mom_pct >= 0,
                       "rgba(29,158,117,0.12)",
                       "rgba(216,90,48,0.12)")
  ) %>%
  add_segments(
    x    = min(index(prices)), xend = max(index(prices)),
    y    = 0, yend = 0,
    name = "Zero line (in/out signal)",
    line = list(color = "#2C2C2A", dash = "dash", width = 1.5)
  ) %>%
  layout(
    title     = paste(params$ticker,
                      "— 12-month rolling return (momentum signal)"),
    xaxis     = list(title = ""),
    yaxis     = list(title = "12-month return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(orientation = "h", x = 0, y = -0.2)
  ) %>%
  config(displayModeBar = FALSE)
```

> When the line is **above zero** the strategy is long — momentum is
> positive. When it dips **below zero** the strategy goes to cash —
> momentum has broken down. Notice how it stayed negative through
> extended bear markets and positive through bull runs.

---

# Results

## Equity curves

```{r equity, fig.height=5}
bh_ret <- dailyReturn(prices)
dates  <- index(prices)

gc_eq  <- cumprod(1 + gc_result$strat_ret)
rsi_eq <- cumprod(1 + rsi_result$strat_ret)
mom_eq <- cumprod(1 + mom_result$strat_ret)
bh_eq  <- cumprod(1 + bh_ret)

plot_ly() %>%
  add_lines(
    x    = dates,
    y    = as.numeric(bh_eq),
    name = "Buy & Hold",
    line = list(color = "#888780", dash = "dash", width = 1.5)
  ) %>%
  add_lines(
    x    = dates,
    y    = as.numeric(gc_eq),
    name = paste0("Golden Cross (MA",
                  params$ma_fast, "/", params$ma_slow, ")"),
    line = list(color = "#185FA5", width = 2.5)
  ) %>%
  add_lines(
    x    = dates,
    y    = as.numeric(rsi_eq),
    name = paste0("RSI (", params$rsi_low,
                  "/", params$rsi_high, ")"),
    line = list(color = "#1D9E75", width = 2.5)
  ) %>%
  add_lines(
    x    = dates,
    y    = as.numeric(mom_eq),
    name = paste0("Momentum (", params$mom_days, "-day)"),
    line = list(color = "#854F0B", width = 2.5)
  ) %>%
  layout(
    title     = paste(params$ticker,
                      "— Equity curves ($1 invested at start)"),
    xaxis     = list(title = ""),
    yaxis     = list(title = "Portfolio value",
                     tickprefix = "$", tickformat = ",.2f"),
    hovermode = "x unified",
    legend    = list(orientation = "h", x = 0, y = -0.2)
  ) %>%
  config(displayModeBar = FALSE)
```

## Drawdown

```{r drawdown, fig.height=4}
calc_dd <- function(ret) {
  eq   <- cumprod(1 + as.numeric(ret))
  peak <- cummax(eq)
  (eq - peak) / peak * 100
}

plot_ly() %>%
  add_lines(
    x = dates, y = calc_dd(bh_ret),
    name = "Buy & Hold",
    line = list(color = "#888780", dash = "dash", width = 1),
    fill = "tozeroy", fillcolor = "rgba(136,135,128,0.07)"
  ) %>%
  add_lines(
    x = dates, y = calc_dd(gc_result$strat_ret),
    name = "Golden Cross",
    line = list(color = "#185FA5", width = 1.5),
    fill = "tozeroy", fillcolor = "rgba(24,95,165,0.09)"
  ) %>%
  add_lines(
    x = dates, y = calc_dd(rsi_result$strat_ret),
    name = "RSI Strategy",
    line = list(color = "#1D9E75", width = 1.5),
    fill = "tozeroy", fillcolor = "rgba(29,158,117,0.09)"
  ) %>%
  add_lines(
    x = dates, y = calc_dd(mom_result$strat_ret),
    name = "Momentum",
    line = list(color = "#854F0B", width = 1.5),
    fill = "tozeroy", fillcolor = "rgba(133,79,11,0.09)"
  ) %>%
  layout(
    title     = paste(params$ticker, "— Drawdown from peak"),
    xaxis     = list(title = ""),
    yaxis     = list(title = "Drawdown (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(orientation = "h", x = 0, y = -0.2)
  ) %>%
  config(displayModeBar = FALSE)
```

## Performance summary

```{r metrics}
calc_metrics <- function(ret, label) {
  ret <- na.omit(as.numeric(ret))
  if (length(ret) == 0 || sum(ret != 0) == 0) return(NULL)

  equity   <- cumprod(1 + ret)
  peak     <- cummax(equity)
  drawdown <- (equity - peak) / peak

  total_ret <- as.numeric(last(equity)) - 1
  ann_vol   <- sd(ret) * sqrt(252)
  ann_ret   <- tryCatch(
    as.numeric(Return.annualized(
      xts(ret, order.by = seq(
        as.Date(params$from),
        by         = "day",
        length.out = length(ret)
      )),
      scale = 252
    )),
    error = function(e) mean(ret) * 252
  )
  sharpe   <- ifelse(ann_vol == 0, 0, round(ann_ret / ann_vol, 2))
  max_dd   <- min(drawdown)
  trading  <- sum(ret != 0)
  win_rate <- ifelse(trading == 0, 0, sum(ret > 0) / trading)
  pct_in   <- mean(ret != 0) * 100

  data.frame(
    Strategy        = label,
    `Total Return`  = paste0(round(total_ret * 100, 1), "%"),
    `Ann. Return`   = paste0(round(ann_ret   * 100, 1), "%"),
    Volatility      = paste0(round(ann_vol   * 100, 1), "%"),
    `Sharpe Ratio`  = sharpe,
    `Max Drawdown`  = paste0(round(max_dd    * 100, 1), "%"),
    `Win Rate`      = paste0(round(win_rate  * 100, 1), "%"),
    `% Time Invested` = paste0(round(pct_in,   1), "%"),
    check.names      = FALSE,
    stringsAsFactors = FALSE
  )
}

metrics <- bind_rows(
  calc_metrics(as.numeric(gc_result$strat_ret),  "Golden Cross"),
  calc_metrics(as.numeric(rsi_result$strat_ret), "RSI Strategy"),
  calc_metrics(as.numeric(mom_result$strat_ret), "Momentum"),
  calc_metrics(as.numeric(bh_ret),               "Buy & Hold")
)

metrics %>%
  kable(align = c("l", rep("r", ncol(metrics) - 1))) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE,
    font_size         = 14
  ) %>%
  row_spec(0, bold = TRUE, background = "#3c8dbc", color = "white") %>%
  row_spec(3, background = "#fff8e8") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(5,
    color = ifelse(
      as.numeric(metrics$`Sharpe Ratio`) >= 1,  "#0F6E56",
      ifelse(
        as.numeric(metrics$`Sharpe Ratio`) >= 0, "#854F0B", "#993C1D"
      )
    ),
    bold = TRUE
  )
```

**Reading the table:**

- **Sharpe Ratio** — green ≥ 1, amber 0–1, red < 0. Higher is better.
- **Max Drawdown** — worst peak-to-trough loss you would have sat through.
- **% Time Invested** — how much of the period each strategy was actually in the market.

---

# Key takeaways

```{r takeaways, results='asis'}
gc_total  <- as.numeric(last(cumprod(1 + gc_result$strat_ret)))  - 1
rsi_total <- as.numeric(last(cumprod(1 + rsi_result$strat_ret))) - 1
mom_total <- as.numeric(last(cumprod(1 + mom_result$strat_ret))) - 1
bh_total  <- as.numeric(last(cumprod(1 + bh_ret)))               - 1

best_val  <- max(gc_total, rsi_total, mom_total, bh_total)
winner    <- dplyr::case_when(
  gc_total  == best_val ~ "Golden Cross",
  rsi_total == best_val ~ "RSI Strategy",
  mom_total == best_val ~ "Momentum",
  TRUE                  ~ "Buy & Hold"
)

gc_dd  <- round(min(calc_dd(gc_result$strat_ret)),  1)
rsi_dd <- round(min(calc_dd(rsi_result$strat_ret)), 1)
mom_dd <- round(min(calc_dd(mom_result$strat_ret)), 1)
bh_dd  <- round(min(calc_dd(bh_ret)),               1)

cat(paste0(
  "- **", winner, "** produced the highest total return ",
  "on **", params$ticker, "** over this period.\n\n",
  "- **Golden Cross** returned **",
  round(gc_total  * 100, 1), "%** with a max drawdown of **",
  gc_dd, "%**.\n\n",
  "- **RSI Strategy** returned **",
  round(rsi_total * 100, 1), "%** with a max drawdown of **",
  rsi_dd, "%**.\n\n",
  "- **Momentum** returned **",
  round(mom_total * 100, 1), "%** with a max drawdown of **",
  mom_dd, "%**. ",
  "It held the position only when the 12-month return was positive, ",
  "sitting in cash otherwise.\n\n",
  "- **Buy & Hold** returned **",
  round(bh_total  * 100, 1), "%** with a max drawdown of **",
  bh_dd, "%** — the passive benchmark.\n\n",
  "> Momentum tends to perform best over long bull markets and ",
  "protects capital during prolonged bear markets by going to cash. ",
  "It underperforms during sharp, fast recoveries where it gets ",
  "back in too slowly.\n\n",
  "> Past performance does not guarantee future results. ",
  "This backtest excludes commissions, slippage, and taxes."
))
```

---

# Methodology

| Item | Detail |
|---|---|
| Data source | Yahoo Finance via `quantmod` — free, no API key |
| Price series | Adjusted close — corrects for dividends and splits |
| Date range | `r params$from` to `r params$to` |
| Golden Cross signal | MA`r params$ma_fast` crosses MA`r params$ma_slow` |
| RSI signal | RSI(`r params$rsi_n`) below `r params$rsi_low` = buy, above `r params$rsi_high` = sell |
| Momentum signal | `r params$mom_days`-day return positive = long, negative = cash |
| Signal lag | All signals lagged 1 trading day — no look-ahead bias |
| Benchmark | Fully invested buy and hold for the entire period |

---

*Generated with R Markdown · `r params$ticker` · `r Sys.Date()`*
